Mirror, mirror on the wall. Which's the best match of them all?

### 2023-12-12 PNAS/index.Rnw
dependency

\dropcap{I}n forensic \ac{evaluations}, a single conclusion often \ac{relies} on  many comparisons, either implicitly or explicitly.
Multiple comparisons arise persistently when developing statistical methods to address scientific problems \citep{benjaminiControllingFalseDiscovery1995}, and can significantly increase the probability of false "discoveries".  Yet, this potentially significant issue is often ignored.  \ac{Now that vast databases and efficient algorithms are routinely used in forensic evaluations to find "matches" to crime scene items, the problems of close non-matches\citep{pcast}, and of the erosion of confidence due to multiple comparisons, become critically important.}
%Researchers have been focusing on building up databases to develop algorithms for objective evidence comparisons.
%\svp{Then, the algorithms facilitate} searches across (vast) databases, reporting the best $N$ matches to a particular piece of evidence.
%This raises the critically important problem of .

@@

A similar concern was raised in the past with regard to DNA \citep{thompson2003} and latent print evaluations  \citep{koehler2021}: one of the issues \citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid train bombing case was that the large size of the IAFIS database used to search for similar prints made it possible to locate `unusually' close non-matches.
\ac{The probability of finding  close non-matches for an item of evidence is directly proportional to the size of the database available for searching;} false "matches" occur more frequently as databases increase in size and algorithms become more powerful. 
% In this paper, we discuss that very close non-matches are far from an unusual phenomenon  but a direct consequence of a large number of comparisons, and simply reflect the power of the search algorithms.

@@

\svp{Compounding this issue,} the use of novel algorithms typically results in a large number of comparisons that may not obvious to the user.
For example, the cross-correlation function\citep{vorburgerApplicationsCrosscorrelationFunctions2011} was one of the first measures proposed for quantifying the similarity  between two patterns in response to the \ac{ 2009 NRC} report \citep{nas2009}, and continues to be used in many pattern searching algorithms to find the best alignment between two images and also to quantify their overall similarity.
Finding the best alignment often consists in sliding one surfaceacross the whole length (for one-dimensional patterns, such as striations) or area (for two dimensional sources, such as impression marks) of the other item while keeping track of the value of a similarity measure.
This mirrors the forensic examination process, where examiners visually rotate or align two surfaces under comparison microscopes.
In order to avoid \ac{problematic} false accusations and a further erosion of the public's trust in science, it is important to address the problem of multiple comparisons by controlling their effect on false discovery rates.

@@

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts \ac{using the putative tool,} which can be compared to the cut surface of the wire.
These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire; depending on the tool design, there can be between two and four cutting surfaces in contact with the substrate.

@@

In order to calculate the number of possible comparisons to be carried out in the course of the examination, we define $\ell$ to be the length of the blade cut, and $d$ to be the diameter of the wire. Without loss of generality, we assume that the wire is fully covered with striations suitable for comparison; if this is not the case, we adjust the value of $d$.
\svp{An illustration of the sliding comparison process is shown in \Cref{fig:wire-blade}.} Imagine that you start from the left end of the blade and move the wire surface \ac{$d$ units at a time} along the blade.
At minimum, there must be $\ell/d$ comparisons in order to assess whether one side of the blade cut matches one side of the wire; in this case, the comparisons are independent of each other (conditional on striation patterns) because \ac{each comparison involves different surfaces.}
%no overlapping data is used.
Of course, this overlooks the possibility of partial matches and assumes perfect alignment between the diameter of the wire and the portion of the blade used to cut the wire, which is unrealistic.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{../fig/wire-blade-comparison-label.png}
\caption{A comparison between a wire and a blade cut requires sliding the wire along the entire blade cut length to determine the best match (or whether there is a match). \svp{Surfaces shown are rendered 2D topographical scans of a wire and blade cut taken with a confocal light microscope.}}\label{fig:wire-blade}
\end{figure}

At a maximum, both the blade and the wire are either digitally scanned at resolution $r$ mm, or visually examined using a microscope with a digital resolution that can be expressed as $r$ equivalent to the digital scan.
In this case, there may be as many as $\ell/r - d/r + 1$ comparisons performed in order to find the optimal alignment between the two samples.
These comparisons are \ac{no longer independent}, as each sequential comparison shares much of the same data with the previous and next comparisons.
% ; such dependency could be expressed via autocorrelation, but we will defer that derivation to a later paper.
\svp{For simplicity}, let us consider the number of comparisons to lie somewhere between \svp{these two estimates}.

\ac{Finally}, we must consider the number of surfaces which must be compared: the wire may have between one and two sets of striae (depending on tool configuration) and there may be between 2 and 4 different blade cut surfaces to examine.
This results in a multiplier of between 2 and 8 comparisons to be conducted to compare the striations on the wire to the striations on the blade cuts.

@@

Let us consider a wire-cutting tool with a 1.5 cm razor blade that meets a cast surface (one such tool is shown in \Cref{fig:tool-pic}); the wire is held against this rectangular cast surface as the blade is pushed into the wire, splitting it in two.
This is a minimal scenario - the wire will acquire striations from one side of the blade, while the blade itself has two cutting edges, which we will call side A and side B.
The blade cuts of a sheet of aluminum that is 1.5 cm in width thus produce two striated edges corresponding to side A and side B which can be compared to other cuts to assess similarity.

>>>>

\dropcap{I}n forensic evaluations, a single conclusion often relies on  many comparisons, either implicitly or explicitly.
Multiple comparisons arise persistently when developing statistical methods to address scientific problems \citep{benjaminiControllingFalseDiscovery1995}, and can significantly increase the probability of false "discoveries".  Yet, this potentially significant issue is often ignored.  Now that vast databases and efficient algorithms are routinely used in forensic evaluations to find "matches" to crime scene items, the problems of close non-matches \citep{pcast}, and of the erosion of confidence due to multiple comparisons, become critically important.
%Researchers have been focusing on building up databases to develop algorithms for objective evidence comparisons.
%\svp{Then, the algorithms facilitate} searches across (vast) databases, reporting the best $N$ matches to a particular piece of evidence.
%This raises the critically important problem of .

@@

These concerns were raised in the past with regard to DNA \citep{thompson2003} and latent print evaluations  \citep{koehler2021}: one of the issues \citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid train bombing case was that the large size of the IAFIS database used to search for similar prints made it possible to locate `unusually' close non-matches.
The probability of finding  close non-matches for an item of evidence is directly linked to the size of the database available for searching: false "matches" occur more frequently as databases increase in size and algorithms are more powerful.
% In this paper, we discuss that very close non-matches are far from an unusual phenomenon  but a direct consequence of a large number of comparisons, and simply reflect the power of the search algorithms.

@@

Compounding this issue, algorithmic results are typically based on a large number of comparisons that may not obvious to the user.
For example, a typical task in the forensic examination process is to find the best alignment between two items of evidence. For one-dimensional patterns, such as striation marks, this consists of sliding one item along the full length of the other item. For two-dimensional patterns, such as impression marks on an area, an additional rotation is involved. In practice, examiners visually rotate or align two surfaces under comparison microscopes. Algorithmically this visual search is mirrored by finding a maximum of the (area) cross-correlation function. The cross-correlation function was one of the first measures proposed \citep{vorburgerApplicationsCrosscorrelationFunctions2011} for quantifying the similarity  between two patterns in response to the  2009 NRC report \citep{nas2009}, and continues to be used in many pattern searching algorithms to find the best alignment between two images and  quantify their overall similarity.

In order to avoid problematic false accusations and a further erosion of the public's trust in science, it is important to address the problem of multiple comparisons by controlling their effect on false discovery rates.

@@

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts using the putative tool which can be compared to the cut surface of the wire.

These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire; depending on the tool design, there can be between two and four cutting surfaces in contact with the substrate.

@@

In order to calculate the number of possible comparisons to be carried out in the course of the examination, we define $\ell$ to be the length of the blade cut, and $d$ to be the diameter of the wire. Here, we assume that striations  suitable for comparison are visible across the full width $d$ of the wire; if this is not the case, we adjust the value of $d$.
An illustration of the sliding comparison process is shown in \Cref{fig:wire-blade}. Imagine that a comparison starts from the left end of the blade and moves the wire surface $d$ units at a time along the blade.
At minimum, there must be $\ell/d$ comparisons in order to assess whether one side of the blade cut matches one side of the wire. We can generally assume that these comparisons are independent  because each comparison involves different surfaces. \hh{We need to further assume that the length $d$ is wide enough to cover some striations; this ensures that the width of $d$ is wider than any significant autocorrelation  observed for the striations on the blade}.
%no overlapping data is used.
\hh{Note, that this is a rather crude assessment of the bare minimal number of comparisons. }% Of course, this overlooks the possibility of partial matches and assumes perfect alignment between the diameter of the wire and the portion of the blade used to cut the wire, which is unrealistic.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{../fig/wire-blade-comparison-label.png}
\caption{A comparison between a wire and a blade cut requires sliding the wire along the entire blade cut length to determine the best match (or whether there is a match). Surfaces shown are rendered 2D topographical scans of a wire and blade cut taken with a confocal light microscope.}\label{fig:wire-blade}
\end{figure}

At a maximum, both the blade and the wire are either digitally scanned at resolution of $r$ pixel per mm, or visually examined using a microscope with a digital resolution that can be expressed as equivalent to the digital scan.
In this case, there may be as many as $\ell/r - d/r + 1$ comparisons performed in order to find the optimal alignment between the two samples.
These comparisons are no longer independent, as each sequential comparison shares much of the same data with the previous and next comparisons.
% ; such dependency could be expressed via autocorrelation, but we will defer that derivation to a later paper.
For simplicity, let us consider the number of comparisons to lie somewhere between these two estimates.

Finally, we must consider the number of surfaces which must be compared: the wire may have between one and two sets of striae (depending on tool configuration) and there may be between 2 and 4 different blade cut surfaces to examine.
This results in a multiplier of between 2 and 8 comparisons to be conducted to compare the striations on the wire to the striations on the blade cuts.

@@

Let us consider a wire-cutting tool with a 1.5 cm razor blade that meets a cast surface (one such tool is shown in \Cref{fig:tool-pic}); the wire is held against this rectangular cast surface as the blade is pushed into the wire, cutting it in two.
This is a minimal scenario -- the wire will acquire striations from one side of the blade, while the blade itself has two cutting edges, which we will call side A and side B.
The blade cuts of a sheet of aluminum that is 1.5 cm in width thus produce two striated edges corresponding to side A and side B which can be compared to other cuts to assess similarity.
